---
title: "Bank Marketing Analysis by Candace Grant"
output: pdf_document
date: "2025-08-31"
---

### Introduction

This data report presents an analysis of a marketing dataset from a Portuguese banking institution's direct marketing campaigns. The dataset focuses on phone-based marketing efforts aimed at promoting term deposits to clients. 

The primary objective is to develop a predictive classification model that determines whether a client will subscribe to a term deposit (binary outcome: 'yes' or 'no'). The campaigns often required multiple contacts with the same client to achieve successful conversions, making this a complex customer behavior prediction problem."

**Dataset Source:**
Moro, S., Rita, P., & Cortez, P. (2014). Bank Marketing [Dataset]. UCI Machine Learning Repository. https://doi.org/10.24432/C5K306.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, 
                      results = 'markup', fig.show = 'asis', fig.keep = 'all', fig.width = 8, fig.height = 6, comment = "")
```
#### Loading packages for data analysis

```{r packages}
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(knitr)
```
#### Import Dataset

```{r}
df <- read_excel("/Users/candace/Downloads/bank+marketing 2/bank/bank-full.xls")

```
#### Dataset Overview and Structure


```{r }
cat("Number of rows:", nrow(df), "\n")
cat("Number of columns:", ncol(df), "\n")

print(names(df))

sapply(df, class)

print(head(df))
```
#### Removing columns, "contact" and "poutcome" where over 70% of values are labeled "unknown."


```{r }
df_clean <- df[, !names(df) %in% c("contact", "poutcome")]
```



```{r }
str(df_clean)
```
#### Renaming columns to be more descriptive

```{r}
library(dplyr)
df_clean <- df_clean %>% rename(
  marital_status = marital,
  education_level = education,
  credit_card_default = default,
  avg_yearly_balance = balance,
  mortgage = housing,
  personal_loan = loan,
  recency = pdays,
  prior_contacts = previous,
  subscribed = y
)
str(df_clean)
```
#### In this section I am creating simple bar plots to give an understanding of the characteristics of potential customers that could shed light on their final decision to subscribe to the term deposit.


```{r}
library(ggplot2)

#### Relationship Plots ####

# Job vs Mortgage
ggplot(df_clean, aes(x = job, fill = mortgage)) +
  geom_bar(position = "fill") +
  labs(title = "Mortgage by Job Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Job vs Personal Loan
ggplot(df_clean, aes(x = job, fill = personal_loan)) +
  geom_bar(position = "fill") +
  labs(title = "Personal Loan by Job Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Side-by-side bars instead of stacked
ggplot(df_clean, aes(x = job, fill = subscribed)) +
  geom_bar(position = "dodge") +
  labs(title = "Term Deposit Subscription by Job Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Conclusion

I began this analysis by examining the dataset structure and discovered an important data quality issue: missing values were coded as 'unknown' rather than traditional NA values, which R does not automatically recognize as missing data.
To address this, I calculated the percentage of 'unknown' values in each column and removed those with greater than 70% missing data, particularly columns that did not contribute to the primary objective of predicting term deposit subscriptions.
Key steps completed:

- Reviewed dataset structure and identified data quality issues
- Renamed columns for better clarity and interpretability
- Removed columns with excessive missing data (>70% 'unknown' values)
- Created visualizations to explore relationships between variables and the target outcome

This preprocessing creates a clean dataset suitable for identifying factors that influence customer decisions to subscribe to term deposits.
